---
title: "A brief introduction to primr"
author: "Pierre Masselot"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{A brief introduction to primr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, include = FALSE}
library(primr)
```

# A basic toy example

Consider the following example, where a response is linearly linked to two
explanatory variables. We create an artifical bump that occurs when
$x_1 > 0.8$ and $x_2 > 0.5$:

```{r, fig.dim = c(7, 7)}
set.seed(12345)
x <- matrix(runif(2000), ncol = 2, dimnames = list(NULL, c("x1", "x2")))
y <- 2 * x[,1] + 5 * x[,2] + 10 * (x[,1] >= .8 & x[,2] >= .5) + rnorm(1000)

plot(x, col = hcl.colors(10)[cut(y, 10)], pch = 16)
```

## Peeling

The initial peeling can be carried out with the function `peeling`. This 
function allows to provide the peeling fraction `alpha`, the stopping
criterion `beta.stop`, as well as the objective function being maximized
and other parameters. See `?peeling`.

```{r}
# Peels the dataset to maximize the box's mean
peel_res <- peeling(y, x)
``` 

## Choosing the best box

The next step is to examine the peeling trajectory in order to choose the
best box. Several possibilities are available for the peeling trajectory. For
instance, the plot below shows both the evolution of the mean and the
relative difference of mean between each step.

```{r, fig.dim = c(7, 10)}
par(mfrow = c(2,1), mar = c(4, 4, 3, 2) + .1)
plot_trajectory(peel_res, type = "b", pch = 16, col = "cornflowerblue", 
  support = 0.11, abline.pars = list(lwd = 2, col = "indianred"), 
  xlab = "")
plot_trajectory(peel_res, type = "b", pch = 16, col = "cornflowerblue",
  support = 0.11, abline.pars = list(lwd = 2, col = "indianred"), 
  ytype = "rel.diff")
``` 

The peeling trajectory shows that the box's mean reaches a plateau 
when the support is lower than 0.11 (identified by the vertical line). 
This is also shown by a large relative difference in the bottom panel.
This value of 0.11 is very close of the theoretical proportion of
observations in the bump created ($0.2 \times 0.5 = 0.1$).  

If we plot the corresponding box on the data, we see that it fits well
the data.

```{r, fig.dim = c(7, 7)}
plot_box(peel_res, pch = 16, ypalette = hcl.colors(10), support = 0.11, 
  box.args = list(lwd = 2))
``` 

## Pasting

The final step is to refine the box's edges by pasting. Let's say we choose the
final box with support equal to 0.01. By pasting, it is possible to slightly
increase the size of the final box, as shown in the figure below.
It means that including the pasted observations increases the box's
average.

```{r, fig.dim = c(7, 7)}
paste_res <- pasting(peel_res, support = 0.01)

plot_box(paste_res, pch = 16, ypalette = hcl.colors(10), npaste = c(0, 2), 
  box.args = list(lwd = 2, border = c("grey", "black"), lty = 1:2))
``` 

# Some special topics

The above example shows the overall framework to perform the PRIM algorithm.
Here, we cover some additional features of the `primr` package.

## Select the box after peeling

The most important part of the PRIM algorithm is the choice of the peeling's
stopping criterion, i.e. which box to keep after the peeling has been
carried out. The `primr` package offers a few tools to help the user in this
choice.

The first one is to analyze the peeling trajectory as was done in the toy
example above. In this example, we chose the highest relative difference of
the trajectory. This can be automatically extracted by the function
`jump.prim`. The example below shows that it allows finding the same box as
before.

```{r, fig.dim = c(7, 7)}
peel_res <- peeling(y, x, beta.stop = 0.05)
chosen <- jump.prim(peel_res)

plot_box(peel_res, pch = 16, ypalette = hcl.colors(10), 
  support = chosen$final.box$support, box.args = list(lwd = 2))
```

Alternatively, the stopping criterion can be chosen through cross-validation
with the function `cv.trajectory`. For instance, on the toy example below,
the cross-validated trajectory also shows the same plateau as before for
support lower than 0.1 precisely. It also
reaches its maximum for the box with support 
equal to 0.032, although there is an important uncertainty. 

```{r, fig.dim = c(7, 7)}
cv_res <- cv.trajectory(y, x)

plot_trajectory(cv_res, type = "b", pch = 16, col = "cornflowerblue", 
  support = 0.1, npeel = which.max(cv_res$yfun), 
  abline.pars = list(lwd = 2, col = "indianred"), 
  xlab = "", xlim = c(0, 0.2), ylim = c(10, 18))
```

Once the peeling has been carried out and the appropriate support chosen,
information about one or several boxes can be retrieved with the function
`extract.box`. For instance, in the following it shows that the box with 
support equal to 0.1 results in limits slightly above the true ones
(0.8 and 0.5).

```{r}
extract.box(peel_res, support = 0.1)
```

## Categorical variables

The `primr` package also supports categorical input variables. For instance,
in the toy example above, we add a third 'letter' variable with a bump
when it takes the value 'b'. 

```{r}
set.seed(54321)
x <- data.frame(x1 = runif(1000), x2 = runif(1000), 
  x3 = sample(letters[1:5], 1000, replace = TRUE))
y <- 2 * x[,1] + 5 * x[,2] + 10 * (x[,1] >= .8 & x[,2] >= .5) + 
  5 * (x[,3] == 'b') + rnorm(1000)
```

Below, we indeed find a box that roughly respects the bump limits, with the
right support of 0.02 ($= 0.2 \times 0.5 \times 0.2$). Note that the 
peeling trajectory shows that the third variable was the first to be peeled
with large chunks of data removed during the first four iterations.

```{r, fig.dim = c(7, 7)}
peel_res <- peeling(y, x)

chosen <- jump.prim(peel_res)
chosen$final.box

plot_trajectory(peel_res, type = "b", pch = 16, col = "cornflowerblue", 
  support = chosen$final.box$support, 
  abline.pars = list(lwd = 2, col = "indianred"), xlab = "")
```

## Directed peeling

Another feature of the `primr` package is the possibility to direct the
peeling in a particular direction.
Let's say that both input variables now induce bumps at both their extremities.

```{r, fig.dim = c(7, 7)}
set.seed(12345)
x <- matrix(runif(2000), ncol = 2, dimnames = list(NULL, c("x1", "x2")))
y <- 10 * (x[,1] <= .2 & x[,2] <= .2) + 10 * (x[,1] >= .8 & x[,2] >= .8) +
  rnorm(1000)

plot(x, col = hcl.colors(10)[cut(y, 10)], pch = 16)
```

We can force the algorithm to peel only the left of the box to work toward 
the bump on the upper right corner.

```{r, fig.dim = c(7, 7)}
peel_left <- peeling(y, x, peeling.side = -1)
chosen <- jump.prim(peel_left)

plot_box(peel_left, pch = 16, ypalette = hcl.colors(10), 
  support = chosen$final.box$support, box.args = list(lwd = 2),
  main = "Left peeling")
```

And conversely, we can direct the peeling to work toward the bump of the 
bottom left corner. Note that different input variables can be directed in 
different directions.

```{r, fig.dim = c(7, 7)}
peel_right <- peeling(y, x, peeling.side = 1)
chosen <- jump.prim(peel_right)

plot_box(peel_right, pch = 16, ypalette = hcl.colors(10), 
  support = chosen$final.box$support, box.args = list(lwd = 2),
  main = "Right peeling")
```

## Alternative objective functions

Although the mean is the most common objective function to maximize, other
functions can be of interest. This can be carried out throught the argument
`obj.fun` in the `peeling` function. For instance, in the following toy 
example, we work toward the high variance area of the data. 

```{r, fig.dim = c(10, 5), out.width = "100%"}
set.seed(1111)
x <- matrix(runif(2000), ncol = 2, dimnames = list(NULL, c("x1", "x2")))
y <- rnorm(1000, sd = 1 + 10 * (x[,1] >= .8 & x[,2] >= .8))

# Standard deviation peeling
peel_res <- peeling(y, x, obj.fun = sd)

par(mfrow = c(1,2))
plot_trajectory(peel_res, type = "b", pch = 16, col = "cornflowerblue", 
  support = 0.036, abline.pars = list(lwd = 2, col = "indianred"), xlab = "",
  xlim = c(0, 0.5))
plot_box(peel_res, pch = 16, ypalette = hcl.colors(10), 
  support = 0.036, box.args = list(lwd = 2))
``` 

The argument `obj.fun` can also be used to minimize the objective function.
The trick is simply to consider the opposite objective function.

```{r, fig.dim = c(7, 7)}
set.seed(3333)
x <- matrix(runif(2000), ncol = 2, dimnames = list(NULL, c("x1", "x2")))
y <- - 10 * (x[,1] <= .2 & x[,2] <= .2) + 10 * (x[,1] >= .8 & x[,2] >= .8) +
  rnorm(1000)

peel_res <- peeling(y, x, obj.fun = function(y) -mean(y))
chosen <- jump.prim(peel_res)

plot_box(peel_res, pch = 16, ypalette = hcl.colors(10), 
  support = chosen$final.box$support, box.args = list(lwd = 2))
```

Finally, more complex objective functions can also be passed to `obj.fun`.
In more complex cases, the function should take two arguments: `y`
representing the response vector, `x` which is the explanatory variable matrix 
and `inbox` which is a boolean
vector indicating the observations inside the current box. 

In the example
below, we use a function to maximize the estimated coefficient of a 
linear regression with a break. The peeling trajectory reaches a 
plateau for supports lower than 0.3.

```{r, fig.dim = c(10, 5), out.width = "100%"}
set.seed(5555)
x <- runif(500)
ym <- 0.5 * x + 5 * (x - 0.7) * (x >= 0.7)
y <- ym + rnorm(500, sd = 0.1)

# Apply to find where the slope is the steepest
peel_res <- peeling(y, x, beta.stop = 0.1, 
  obj.fun = function(y, x, inbox){
    dat <- data.frame(y, x)
    coef(lm(y ~ x, data = dat[inbox,]))[2]
})

par(mfrow = c(1,2))
plot_trajectory(peel_res, type = "b", pch = 16, col = "cornflowerblue", 
  support = 0.3, abline.pars = list(lwd = 2, col = "indianred"))
plot_box(peel_res, pch = 16, ypalette = hcl.colors(10), 
  support = 0.3, box.args = list(lwd = 2))
lines(sort(x), ym[order(x)], col = "red", lwd = 2)

```
